Bayesian phylogenetic estimation of fossil ages

### 2015-09-03 fossilDating.tex
small edits

% 250 words
\abstract{
Recent advances have allowed for both morphological fossil evidence and molecular sequence data to be integrated into a single combined inference of divergence dates under the rule of Bayesian probability. 
In particular the fossilized birth-death tree prior and the Lewis-MK model of evolution of discrete morphological change allow for the estimation of both divergence times and the phylogenetic relationships between fossil and extant taxa. 
We exploit this statistical framework to investigate the internal consistency of these models by estimating the phylogenetic age of each fossil in turn, within two rich and well-characterized data sets of fossil and extant species.
We find that we can accurately estimate the age of individual fossils based only on phylogenetic evidence. 
In fact in the two data sets we analyze the {\em phylogenetic age} of a fossil species is on average only $\sim 2$ My from the midpoint age of the geological strata from which the fossil species was excavated. 
The high level of internal consistency found in our analyses provides strong evidence that the Bayesian statistical model employed is a good fit for both the geological and morphological data, and provides striking evidence from real data that the framework used can accurately model the evolution of discrete morphological traits coded from fossil and extant taxa. 
We anticipate that this approach will have diverse applications including estimating dates for fossils that are temporally unconstrained, testing the ``morphological clock'', and as part of evaluating model assumptions when controversial phylogenetic hypotheses are obtained based on combined divergence dating analyses.
}

@@

Traditionally the practice has been to use one or more fossils as ``node calibrations'' by associated their geologically-derived age to a particular divergence in a molecular phylogeny. The age of the fossil is determined either by radiometric aging of strata above and/or below the strata, or more commonly by biostratigraphy. The difficulty lies in determining the appropriate ancestral species in the molecular phylogeny to associate the fossil with and the details of how this should be achieved. Once achieved, node calibration confers age estimates to the remaining ancestral divergences in the phylogenetic tree by the assumption of a strict or relaxed molecular clock \autocite{Thorne1998,thorne2005,yang2006,Drummond2006}.

@@

We extend this work into the realm of morphological evolution by presenting a statistical model of evolution that generates an expectation on the distribution of fossils and their morphological characters. 
This model has been previously presented in the context of divergence time dating \autocite{gavr2014,gavryushkina2015bayesian}. 
It is distinct from alternative divergence time dating approaches in that it provides an explicit treatment of the temporal information contained in fossil remains, whether or not related molecular sequence data is available. 
This leads to an estimate of the age of the most recent common ancestor of a group of fossils. 
We exploit this framework to attempt the estimation of the phylogenetic age of individual fossils based solely on morphological data and their phylogenetic affinities to related fossils of known age. 
The method is applied to two rich and well-characterized morphological data sets: (i) extant penguins and 36 fossil relatives \autocite{ksepka2010,ksepka2012}, (ii) a sample of nine extant canids and 123 fossil relatives \autocite{Slater2015}.

@@

The probabilities $p(H_1|D,M)$ and $p(H_2|D,M)$ are obtained directly from the MCMC output.
It remains to calculate the probabilities $p(H_1|M_\mathcal{T})$ and $p(H_2|M_\mathcal{T})$.
In our case, the model $M_\mathcal{T}$ is the fossilized birth-death process with the specified priors on its parameters, in our chosen parameteriztion the time of the start of the process $T$, the net diversification rate $d$ (= speciation rate - extinction rate), the turnover $r$ (= extinction rate / speciation rate) and the sampling probability $s$ (=sampling rate / (extinction rate + sampling rate)).
We derive the probability density of sampling a fossil at time $t$ in the past, given the model $M_\mathcal{T}$. 
For a given $T$, $d$, $r$ and $s$, the probability density of sampling a fossil at time $t$, given the process does not go extinct for time $T$, is,
$$p(t|T,d,r,s) = \frac{1}{1-p_0(T;d,r)} \sum_{k=1}^\infty k \psi p_k(T-t;d,r) (1-p_0(t;d,r)^k)$$
with $\psi=\frac{s}{1-s} \frac{rd}{1-r}$ being the sampling rate, and $p_i(t;d,r)$ being the probabiltiy of a single lineage producing $i$ surviving lineages at time $t$.
We simplify,
\begin{eqnarray*}
p(t|T,d,r,s) &=& \frac{1}{1-p_0(T;d,r)}  \psi p_1(T-t|d,r) \\ & & \times \left[  \frac{1}{(1-p_0(T-t|d,r)/r)^2} - \frac{p_0(t|d,r)}{(1-p_0(t|d,r) p_0(T-t|d,r)/r)^2}  \right]\\
&=&  \psi  \left[ \frac{e^{d(T-t)}}{1-p_0(T;d,r)} - \frac{p_0(T|d,r)-p_0(T-t|d,r)}{(1-p_0(T-t|d,r))(1-p_0(t|d,r))} \right]
\end{eqnarray*}

@@

This simple exercise highlights that whenever using Bayes factor to test a hypothesis, we have to first investigate what our prior on the hypothesis is. In our example, the first prior seemed reasonable for the parameters specified, however by allowing very large diversification rates, the resulting huge trees dominate the shape of our prior for our hypothesis  $p(t|M_\mathcal{T})$, namely most samples are very very recent.  Thus this prior  puts a negligible weight on hypothesis $H_1$ for older fossils.
CHECK IF THIS IS TRUE: Changing to our new prior has immense impact on the Bayes factor analysis, but in our case has a minor effect on the posterior distribution of trees / parameters.

>>>>

% 250 words
\abstract{
Recent advances have allowed for both morphological fossil evidence and molecular sequences to be integrated into a single combined inference of divergence dates under the rule of Bayesian probability. 
In particular the fossilized birth-death tree prior and the Lewis-MK model of evolution of discrete morphological change allow for the estimation of both divergence times and phylogenetic relationships between fossil and extant taxa. 
We exploit this statistical framework to investigate the internal consistency of these models by estimating the phylogenetic age of each fossil in turn, within two rich and well-characterized data sets of fossil and extant species.
We find that we can accurately estimate the age of individual fossils based only on phylogenetic evidence. 
In fact in the two data sets we analyze the {\em phylogenetic age} of a fossil species is on average only $\sim 2$ My from the midpoint age of the geological strata from which it was excavated. 
The high level of internal consistency found in our analyses provides strong evidence that the Bayesian statistical model employed is a good fit for both the geological and morphological data, and provides striking evidence from real data that the framework used can accurately model the evolution of discrete morphological traits coded from fossil and extant taxa. 
We anticipate that this approach will have diverse applications beyond divergence time dating, including dating fossils that are temporally unconstrained, testing the ``morphological clock'', and for evaluating model assumptions when controversial phylogenetic hypotheses are obtained based on combined divergence dating analyses.
}

@@

Traditionally the practice has been to use one or more fossils as ``node calibrations'' by associated their geologically-derived age to a particular divergence in a molecular phylogeny. The age of the fossil is determined either by radiometric aging of strata above and/or below the strata, or more commonly by biostratigraphy. 
The difficulty lies in determining the appropriate ancestral species in the molecular phylogeny to associate the fossil with and the details of how this should be achieved within a full statistical inference framework \autocite{Ho2009, Heled2012, HeledDrummond2015}.
Once achieved, node calibration confers age estimates to the remaining ancestral divergences in the phylogenetic tree by the assumption of a strict or relaxed molecular clock \autocite{Thorne1998,thorne2005,yang2006,Drummond2006,drummond2010}.

@@

We extend this approach into the realm of discrete morphological evolution by presenting a statistical model of evolution that generates an expectation on the distribution of fossils and their morphological characters. 
This model has been previously presented in the context of divergence time dating \autocite{gavr2014,gavryushkina2015bayesian}. 
It is distinct from alternative divergence time dating approaches in that it provides an explicit treatment of the temporal information contained in fossil remains, whether or not related molecular sequence data is available. 
This leads to an estimate of the age of the most recent common ancestor of a group of fossils. 
We exploit this framework to attempt the estimation of the phylogenetic age of individual fossils based solely on morphological data and their phylogenetic affinities to related fossils of known age. 
The method is applied to two rich and well-characterized morphological data sets: (i) extant penguins and 36 fossil relatives \autocite{ksepka2010,ksepka2012}, (ii) a sample of nine extant canids and 123 fossil relatives \autocite{Slater2015}.

@@

The probabilities $p(H_1|D,M)$ and $p(H_2|D,M)$ are obtained directly from the MCMC output.
It remains to calculate the probabilities $p(H_1|M_\mathcal{T})$ and $p(H_2|M_\mathcal{T})$.
In our case, the model $M_\mathcal{T}$ is the fossilized birth-death process with the specified priors on its parameters, in our chosen parameteriztion the time of the start of the process $T$, the net diversification rate $d$ (= speciation rate - extinction rate), the turnover $r$ (= extinction rate / speciation rate) and the sampling probability $s$ (=sampling rate / (extinction rate + sampling rate)).
We derive the probability density of sampling a fossil at time $t$ in the past, given the model $M_\mathcal{T}$. 
For a given $T$, $d$, $r$ and $s$, the probability density of sampling a fossil at time $t$, given the process does not go extinct for time $T$, is,
$$p(t|T,d,r,s) = \frac{1}{1-p_0(T;d,r)} \sum_{k=1}^\infty k \psi p_k(T-t;d,r) (1-p_0(t;d,r)^k)$$
with $\psi=\frac{s}{1-s} \frac{rd}{1-r}$ being the sampling rate, and $p_i(t;d,r)$ being the probability of a single lineage producing $i$ surviving lineages at time $t$.
We simplify,
\begin{eqnarray*}
p(t|T,d,r,s) &=& \frac{1}{1-p_0(T;d,r)}  \psi p_1(T-t|d,r) \\ & & \times \left[  \frac{1}{(1-p_0(T-t|d,r)/r)^2} - \frac{p_0(t|d,r)}{(1-p_0(t|d,r) p_0(T-t|d,r)/r)^2}  \right]\\
&=&  \psi  \left[ \frac{e^{d(T-t)}}{1-p_0(T;d,r)} - \frac{p_0(T|d,r)-p_0(T-t|d,r)}{(1-p_0(T-t|d,r))(1-p_0(t|d,r))} \right]
\end{eqnarray*}

@@

This simple exercise highlights that whenever using Bayes factor to test a hypothesis, we have to first investigate what our prior on the hypothesis is. 
In our example, the first prior seemed reasonable for the parameters specified, however by allowing very large diversification rates, the resulting huge trees dominate the shape of our prior for our hypothesis  $p(t|M_\mathcal{T})$, namely most samples are very very recent.  Thus this prior  puts a negligible weight on hypothesis $H_1$ for older fossils.
\todo{CHECK IF THIS IS TRUE:} Changing to our new prior has immense impact on the Bayes factor analysis, but in our case has a minor effect on the posterior distribution of trees / parameters.
